### 2025-08-29 docs/thesis_manuscript/manuscript.tex
including the dynamics analysis

Understanding this facet of evolution is [difficult]. Unlike with [allelic mutation], genome evolution occurs up to several orders of magnitude less freqently, and is more heterogenous in its manifestation. We define here genome evolution as a mutational process resulting in either the expansion or reduction in the size of the [functional/coding] genome, stemming from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors, transposon activity, or whole genome duplications, as well as genes arising from noncoding sequence through de novo gene birth, while genome-reducing mechanisms are pseudogenization and deletions via various replication errors. here are some more specifics about the expansion/reduction process. here is why they are difficult to observe or create experiments to test. here is why is would be beneficial to create models that allow us to ask [] questions that bridge the gap between theory and experiment. 
Owing to a large protein fold space, evolution has generally favored mechanisms of duplication and neofunctionalization amongst new genes through horizontal gene transfer, transposon activity, or whole-genome duplications. (by doing so the cell acheives two things: 1. relieve pleiotropic effects, 2. lessen time to navigate sequence space). Alternatively, unused regions of the geome may also see genes spontaneously arise. (this process is much slower and harder to observe). In either case, these new genes are expected to accumulate neutrally, and over time have the opportunity to gain beneficial functionality. 
Although organisms are effective at removing unnecessary genes through pseudogenization, (this is a slow process, and still does not account for historical contingencies)
Different organisms may happen to adapt to the same environment in different way

@@

As has been noted by studies of the \textit{de novo} evolution of genes, their emergence out of unused genetic substrate remains difficult to observe experimentally. (why? duplication vs de novo navigation of landscapes). There is thus a usefulness to generation of models that allow us to form hypotheses about these difficult to test systems.

>>>>

Understanding this facet of evolution is [difficult]. Unlike with [allelic mutation], genome evolution occurs up to several orders of magnitude less freqently, and is more heterogenous in its manifestation. We define here genome evolution as a mutational process resulting in either the expansion or reduction in the size of the [functional/coding] genome, arising from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors, transposon activity, or whole genome duplications, as well as genes arising from noncoding sequence through de novo gene birth, while genome-reducing mechanisms include pseudogenization and deletions via various types of replication errors. Evolution has been observed to utilize both genome reduction and expansion spur adaptation to increase the organism's \textit{fitness}, conceptualized as some metric representing success e.g. growth rate within an environment. However, some of the more impactful processes such as neofunctionalization of duplications, de novo gene birth, and WGD, remain difficult to experimentally induce and test. (why?) There is thus a usefulness to the creation of mathematical models that bridge the gap between theory and experiment and allow us to form hypotheses about these phenomena and their roles in long-term evolution.\\

Given a model of genome size evolution, what are we interested in testing? (path dependency)\\

here are some more specifics about the expansion/reduction process. here is why they are difficult to observe or create experiments to test.
Owing to a large protein fold space, evolution has generally favored mechanisms of duplication and neofunctionalization amongst new genes through horizontal gene transfer, transposon activity, or whole-genome duplications. (by doing so the cell acheives two things: 1. relieve pleiotropic effects, 2. lessen time to navigate sequence space). Alternatively, unused regions of the geome may also see genes spontaneously arise. (this process is much slower and harder to observe). In either case, these new genes are expected to accumulate neutrally, and over time have the opportunity to gain beneficial functionality. 
Although organisms are effective at removing unnecessary genes through pseudogenization, (this is a slow process, and still does not account for historical contingencies)
Different organisms may happen to adapt to the same environment in different way

@@

As has been noted by studies of the \textit{de novo} evolution of genes, their emergence out of unused genetic substrate remains difficult to observe experimentally. (why? duplication vs de novo navigation of landscapes).
